---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# PoDCall

<!-- badges: start -->
<!-- badges: end -->

# Introduction
PoDCall (Positive Droplet Caller) 
[M. Jeanmougin et al](https://doi.org/10.1093/bioinformatics/btac766) is a 
package that aims to provide a robust calling of positive droplets in DNA 
methylation droplet digital PCR (ddPCR) experiments performed on the Bio-Rad 
platform.
PoDCall provides functions that reads files exported from QuantaSoft or 
QX Manager containing amplitudes from a run of ddPCR (one 96 well plate), sets 
thresholds for both channels of each individual well and calculates 
concentrations and normalized concentration for each well.The resulting 
threshold table can optionally be written to file automatically by the main 
workflow function. PoDCall also offers functionality for plotting, both 
individual wells and multiple well plots. Plots for individual wells can be made
and saved as .pdf-files as part of the main workflow function, or by calling the
various plotting functions individually. 

## Gaussian Mixture Models
DdPCR experiments generate a mixture of droplets, positive droplets which 
contain the target that will be amplified, and negative droplets that does not 
contain the target and show no amplification. PoDCall relies on fitting Gaussian
mixture models to set thresholds for each individual well 
that will be used to classify the droplets as either positive or negative.
For more details on the concepts of PoDCall, see the application note.


## Input Data
The input data is .csv-files exported from 'QuantaSoft' or 'QX Manager, and each
file contains the amplitude values of droplets from one well of a 96 well plate.
The first two columns of the files should have headers 'Ch1 Amplitude' and 
'Ch2 Amplitude'. To read in data, use the function importAmplitudeData, which 
will read all amplitude files in the directory given as argument. Each file will
be stored as a named data frame in a list, where the name will be the well ID. 
For this reason, all raw data files in the directory given as argument should 
be from the same 96 well plate to avoid well coordinate duplicates.


# Installation
PoDCall requires some packages to be installed, and if any required packages are
not yet installed, the installation of PoDCall should take care of it (you will 
be prompted to install the packages that are missing).   

The released version of PoDCall can not yet be installed from 
[BIOCONDUCTOR](http://bioconductor.org/), but for now it can be installed 
from GitHub or from a source file.

``` {r installation, eval=FALSE}
## Install PoDCall from Bioconductor
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("PoDCall")

## Install PoDCall from GitHub
install.packages("devtools")
devtools::install_github("HansPetterBrodal/PoDCall")

## Install PoDCall from source file
install.packages("remotes")
remotes::install_local("path/to/PoDCall_0.99.0.tar.gz")

```
After installing PoDCall and the required packages, PoDCall can be loaded with:

``` {r loading, eval=TRUE}

library(PoDCall)

```

# Example / Usage

One step of setting thresholds includes a random sampling of droplets to greatly
reduce running time. To ensure reproducible results it is recommended to set a
seed using `set.seed()`.
To run the full PoDCall workflow, call the function `podcallDdpcr()`:

``` {r, eval=FALSE}
## Run PoDCall
thresholdTable <- podcallDdpcr(dataDirectory="path/to/data/", 
                                software="QuantaSoft")
```
Where "path/to/data/" is the path of the directory that contains amplitude files
from a well plate, in which the files have names that end with 
"_wellID_amplitude.csv"."software" is the software that was used to export the 
data (amplitude) files and the sample sheet. Must be either "QuantaSoft" or 
"QX Manager". Since the different software versions format amplitude files and
sample sheet differently, it is important to set the correct value as argument 
to ensure that data and sample sheet is read correctly.


## Optional arguments
The following arguments have default values, but can be given other values if
desired. For example to write results to file, which is disabled by default.

### sampleSheetFile 
Path to sample sheet file. Must be a .csv file exported from QuantaSoft and must
include the following columns: Well, Sample, TargetType and Target. The entries
in the column TargetType must be either 'Ch1Unknown' or Ch2Unknown, and is used
to extract rows with information from either channel 1 or channel 2. An example
file has been included in the package, which can be found using 
`system.file("extdata", "Sample_names.csv", package="PoDCall")`

### B 
The number of permutations used by the likelihood ratio test (LRT) 
which decides the number of components in the model fitted from the data. 
Default value `B=200`.

### Q 
A parameter used for calling outliers. Q is multiplied with the interquartile 
range of the distribution of amplitude values to determine if droplets of 
extreme amplitude values are to be considered outliers. The default value is 
`Q=9`, which has been determined through cell line experiments and testing.
A higher Q will generally result in a higher or more strict threshold. Q 
provides an option to adjust how thresholds are set in a systematic and 
reproducible way. It is recommended to try a few different values and visually
inspect the results.

### refwell 
The well used as reference when calculating the shift in baseline
between wells. By default `refwell=1`, but can be changed in cases where the 
first well is not suited to be used.  

### ch2 
If channel 2 is not in use, set ch2 = FALSE to avoid error caused by 
empty channel 2 column. Default is `ch2=TRUE`.

### resultsToFile 
The user can choose to let PoDCall save the results table as a .csv-file by 
setting `resultsToFile=TRUE` (default: `resultsToFile = FALSE`). When 
resultsToFile is set to TRUE, a results directory will be created where the
result file will be saved. The results directory will have the same name as the
data directory with "_results" added: "path/to/data_results/

### plots 
The user can choose to make plots that are written to file by setting 
`plots=TRUE` (default: `plots=FALSE`). Plots will be saved to the results 
directory created when `resultsToFile=TRUE`. The results directory will also be
created if only `plots=TRUE`.

### resPath
Optional argument to specify a directory for writing results file(s) to other 
than the results directory created by default. Requires `resultsToFile=TRUE`.


## Threshold table columns
The table that is returned when running `podcall_ddpcr()` contains columns with 
more or less self-explanatory column names, and well ID (well coordinates) as
rownames:

### sample_id
If a sample sheet file is provided, this will have the sample ID from the sample
sheet. Otherwise empty

### thr_target 
the threshold set for channel 1, assumed to be the target

### thr_ctrl 
The threshold set for channel 2, assumed to be the control

### pos_dr_target
The number of positive droplets in channel 1 (target)

### pos_dr_ctrl 
The number of positive droplets in channel 2 (control)

### tot_droplets 
Number of droplets.

### c_target 
Concentration of target, calculated by the formula 
$-\log\dfrac{\dfrac{\text{neg_drop_tar}}{\text{tot_droplets}}}{0.000851}$ 
where neg_drop_tar is number of negative droplets in channel 1 (target).

### c_ctrl 
Concentration of control, calculated by the formula 
$-\log\dfrac{\dfrac{\text{neg_drop_ctrl}}{\text{tot_droplets}}}{0.000851}$ 
where neg_drop_ctrl is number of negative droplets in channel 2 (control).

### c_norm_4Plex 
Normalized concentration with 4Plex as control, calculated by the formula 
$\dfrac{\text{c_target}}{\text{c_ctrl}}\cdot400$

### c_norm_sg 
Normalized concentration with single gene as control, calculated by the formula 
$\dfrac{\text{c_target}}{\text{c_ctrl}}\cdot100$

### q
The value used for Q

### target_assay
The assay used for target channel, provided via sample sheet.

### ctrl_assay
The assay used for control channel, provided via sample sheet.

### ref_well
The well used as reference well for setting threshold.

# PoDCall functions

`podcallDdpcr()` is the main wrapper function that returns a table with the 
results of PoDCall to the user. This function uses a set of functions that read 
the amplitude data from file, set thresholds and make plots. All functions 
involved can be used individually should the user only want to use some of the 
functionality of PoDCall. Also see help files for more details about the 
functions and their arguments.

## `importAmplitudeData()` 
Reads .csv-files with amplitude data outputted from QuantaSoft or QX Manager and
store the data in a list, one data frame per well. Each element in the list will
be named using it's well ID (coordinate) of the 96 well plate that the sample 
belong to.

```{r import-amplitude-data, echo=TRUE}
## Path to example data files included in PoDCall
path <- system.file("extdata", "Amplitudes/", package="PoDCall")

## Read in data files
dataList <- importAmplitudeData(dataDirectory=path, skipLines=0)
str(dataList)
```

## `importSampleSheet()` 
Reads a .csv-file outputted from QuantaSoft or QX Manager to get information 
about the samples: Sample name/id, Assay for target and control. 

```{r import-sample-sheet, echo=TRUE}
## Path to example files included in PoDCall
path <- system.file("extdata", "Sample_names.csv", package="PoDCall")

## Select wells to get information for
well_id <- c("A04", "B04", "D04")

## Read in sample sheet information for selected wells
sampleSheet <- importSampleSheet(sampleSheet=path, well_id=well_id, 
                                software="QuantaSoft")
print(sampleSheet)

```

## `podcallThresholds()` 
Takes a list of data frames, one for each well, as argument and sets 
individual thresholds for each channel of each well. It returns a table with 
thresholds, number of positive droplets, concentrations etc. The number of 
permutations for likelihood ratio test is by default set to `B=400` as a 
compromise between run time and stability of the results. The parameter for 
calling outliers is by default set to `Q=9`. Higher Q means more conservative 
(higher) thresholds, lower Q will result in over all lower thresholds.

```{r set-thresholds, eval=FALSE}
## Path to example data files included in PoDCall
path <- system.file("extdata", "Amplitudes/", package="PoDCall")

## Read in data files
ampData <- importAmplitudeData(dataDirectory=path, skipLines=0)

## Calculate thresholds, metrics, concentrations
thresholdTable <- podcallThresholds(plateData=ampData)
print(thresholdTable)
```

## `podcallChannelPlot()`
Takes the threshold and amplitude values corresponding to a channel of a well as
arguments, calls functions that makes scatter plot and histogram and draws a 
plot with both.

```{r channel-plot, echo=TRUE}
## Read in data and threshold table
path <- system.file("extdata", "Amplitudes/", package="PoDCall")
ampData <- importAmplitudeData(dataDirectory=path, skipLines=0)
data("thrTable")
thresholdTable <- thrTable

## Select channel and well to plot
ch <- 1 # target channel
well_id <- names(ampData)[1] # First well in list

## Plot title
plotTitle <- paste0(well_id, ", Ch", ch)

## Create plot
podcallChannelPlot(channelData=ampData[[well_id]][,ch],
                    thr=thresholdTable[well_id, "thr_target"],
                    channel=ch,
                    plotId=plotTitle)
```

## `podcallScatterplot()` 
Takes the threshold and amplitude values corresponding
to a channel of a well as argument and returns a scatter plot.

```{r scatter-plot, echo=TRUE}
## Read in data and threshold table
path <- system.file("extdata", "Amplitudes/", package="PoDCall")
ampData <- importAmplitudeData(dataDirectory=path, skipLines=0)
thresholdTable <- thrTable

## Select channel and well to plot
ch <- 1 # target channel
well_id <- names(ampData)[1] # First well in list

## Plot title
plotTitle <- paste0(well_id, ", Ch", ch)

## Create plot
podcallScatterplot(channelData=ampData[[well_id]][,ch],
                    thr=thresholdTable[well_id, "thr_target"],
                    channel=ch,
                    plotId=plotTitle)

```

## `podcallHistogram()` 
Takes the threshold and amplitude values corresponding to
a channel of a well as argument, and returns a histogram.

```{r plot-histogram, echo=TRUE}
## Read in data and threshold table
path <- system.file("extdata", "Amplitudes/", package="PoDCall")
ampData <- importAmplitudeData(dataDirectory=path, skipLines=0)
thresholdTable <- thrTable

## Select channel and well to plot
ch <- 1 # target channel
well_id <- names(ampData)[1] # First well in list

## Plot title
plotTitle <- paste0(well_id, ", Ch", ch)

## Create plot
podcallHistogram(channelData=ampData[[well_id]][,ch],
                thr=thresholdTable[well_id, "thr_target"],
                channel=ch,
                plotId=plotTitle)
```

## `podcallMultiplot()` 
takes a list of data frames with amplitude data, one per well, and their 
respective thresholds as arguments and returns faceted scatter plots suitable 
for comparing wells.

```{r comparison-plot, echo=TRUE}
## Read in data and threshold table
path <- system.file("extdata", "Amplitudes/", package="PoDCall")
ampData <- importAmplitudeData(dataDirectory=path, skipLines=0)
thresholdTable <- thrTable

## Channel to plot
ch <- 1

## Create comparison plot
podcallMultiplot(plateData=ampData,
                thresholds=thresholdTable[names(ampData),], 
                channel=ch)

```

# PoDCall shiny application

PoDCall does also include an application powered by shiny that launches in a 
web browser. The application provides a user friendly and interactive interface 
to the functionality of PoDCall. To start the app:

```{r launch-shiny-app, eval=FALSE}
podcallShiny()
```

# PodCall example data
There are some amplitude files and a sample sheet included in the package that 
are intended to be used to run examples and to try out the functionality of 
PoDCall. The data files are from a real experiment performed with cell line 
samples. There is also a threshold table computed from the example data 
included. PoDCall takes a few minutes to run due to bootstrapping, and this 
table is used in examples for functions where threshold is an argument.

## Cell Line Amplitude Data
The cell line amplitude data files can be found in the "extdata" subdirectory 
of the package directory and can be found using `system.file()`:
```{r example-data, eval=TRUE}
## Path to files
path <- system.file("extdata", "Amplitudes/", package="PoDCall")

## List files
list.files(path)
```
The control assay used for the samples in the example data files is an assay
developed in-house called 4Plex
[H. Pharo et al](http://dx.doi.org/10.1186/s13148-018-0456-5).

## Calculated Threshold Table
The already calculated threshold table is instantly available when PoDCall is
loaded, and is available as an object called `thrTable`. See `?thrTable` for 
help file with documentation on the table. 

```{r example-thresholds, echo=TRUE}
## The threshold table
thrTable
```

# Session info
Here is the output of `sessionInfo()` on the system on which this document was
compiled

```{r session-info, eval=TRUE}
sessionInfo()
```



